Fast time-domain electromagnetic modelling in the frequency domain

### 2020-10-15 ManuscriptTime.tex
Update figures within revision

\address{
\footnotemark[1]TU Delft,
Building 23,
Stevinweg 1 / PO-box 5048,
2628 CN Delft;\\
\footnotemark[2]Shell Global Solutions International BV,
Grasweg 31, 1031 HW Amsterdam;\\
E-mail: \href{mailto:Dieter@Werthmuller.org}{Dieter@Werthmuller.org};\\
\textbf{Keywords}: Fourier transforms, CSEM, 3D modelling.
}

@@

%%fakesection ===    ABSTRACT    ===
\begin{abstract}
%
Modelling three-dimensional time-domain controlled-source electromagnetic data
with frequency-domain solvers is an alternative to time-domain solvers. It
requires the computation of many frequencies if standard Fourier transforms are
used, which can make it prohibitively expensive in comparison with a
time-domain computation. The speed of time-domain modelling with
frequency-domain computation is defined by three key points: solver, method and
implementation of the Fourier transformation, and gridding.
%
The faster the frequency-domain solver, the faster time-domain modelling will
be. It is important that the solver is robust over a wide range of frequencies.
A three-dimensional, iterative, frequency-domain multigrid solver satisfies
this requirement, but any other robust solver will do.
%
The method should require as few frequencies as possible while remaining
robust. As the frequency range spans many orders of magnitude, the required
frequencies are ideally equally spaced on a logarithmic scale. The fast
frequency-to-time domain method uses either the digital linear filter method
or the logarithmic fast Fourier transform together with a careful selection of
frequencies and interpolation. This methodology requires typically 15 to 20
frequencies to cover a wide range of offsets in one-dimensional test models.
%
The gridding should be frequency-dependent, which is accomplished by making it
a function of skin depth. Optimising for the least number of required cells for
a given frequency should be combined with optimising for computational speed.
%
Looking carefully at these three points results in much smaller computation
times with speedup factors of ten or more over previous results. A
frequency-domain computation of time-domain data can therefore be competitive
to time-domain solvers if the required frequencies and the corresponding grids
are carefully chosen.
%
\end{abstract}

@@

Being able to model CSEM data both in the frequency-domain and in the
time-domain can be desirable, as both domains have advantages and
disadvantages. One way to achieve this is to implement Maxwell's equation in
both domains, as it is done, for instance, in \simpeg \citep{CAG.15.Cockett}.
Another approach is to have Maxwell's equation only implemented in one domain,
and use Fourier transforms to go to the other. However, this approach can be
costly, as many frequencies over a wide range are required to go from the time
domain to the frequency domain, or many times over a wide range for the
opposite direction. We present a methodology which significantly reduces the
amount and range of the required frequencies, translating into a significant
reduction in computation time. The required computation grids for low
frequencies (in our case in the order of 0.001\,Hz) and high frequencies
(around 100\,Hz) are hugely different. Low frequencies can be computed on a
coarser grid, but they require a much larger domain in order to avoid boundary
effects. High frequencies, on the other hand, require denser gridding, but they
are much more limited in reach. An adaptive gridding scheme is therefore
required. We build our approach upon \cite{GEO.07.Plessix}, who presented such
an adaptive gridding for multi-frequency (and multi-source) CSEM modelling.

@@

For the numerical computations we use the open-source (Apache License 2.0) code
\emg3d \citep{JOSS.19.Werthmuller}, a multigrid solver based on
\cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov subspace
solvers or as a solver on its own. The multigrid approach works fine for the
diffusive approximation of Maxwell's equation, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which \emg3d has implemented known improvements such as line-relaxation and
semicoarsening \citep{ECCFD.06.Jonsthovel}, with a non-standard Cholesky
decomposition to speed up the computation \citep{GEO.08.Mulder}. One of the big
advantages of the multigrid method is that it scales linearly (optimal) with
the grid size in both CPU and RAM usage \citep{B.Springer.20.Mulder}. This
makes it feasible to run even big models on standard computers, without the
need for big clusters. All the examples in this article are run on a laptop
with an i7-6600U\,CPU\,@\,2.6\,GHz (x4) and 16\,GB of memory, using Ubuntu
18.04 and Python 3.7. For the semi-analytical computations of layered models we
use the open-source (Apache License 2.0) code \empymod
\citep{GEO.17.Werthmuller}.

@@

There are three key-parameters that have to be defined: Minimum and maximum
frequency ($f_\mr{min}, f_\mr{max}$) and the number of frequencies per decade
($n_\mr{dec}^f$). Trial-and-error with a 3D code is very time-intensive.
However, a simplified, layered model for the required survey setup and a fast
1D modeller makes it possible to estimate these parameters easily. To this end,
we created a graphical, interactive tool using \empymod, as shown in
Figure~\ref{fig:GUI-FFTLog-lin}.
%
\plot{GUI-FFTLog-lin}{width=\tcwidth}{
  Interactive frequency selection for a user-provided layered model (the shown
  model parameters are described in the text). This example shows the impulse
  response at an offset of 5\,km, for which it uses FFTLog from 0.001\,Hz to
  10\,Hz with five frequencies per decade. Note that the error is clipped for
  values smaller than 0.01\,\% and bigger than 100\,\%.
}
%
The example model is a marine scenario with 1\,km water depth of resistivity
$\rho=0.3\,\Omega$\,m ($\rho = \sigma^{-1}$) and a 100\,m thick target of
100\,$\Omega$\,m at 1\,km below the seafloor in a background of 1\,$\Omega$\,m.
The source is 50\,m above the seafloor, the receiver is on the seafloor, and
the response is the inline $x$-directed $E$-field. The left subplot shows the
imaginary part in the frequency domain and the right plot the corresponding
impulse response in the time domain. The red lines are the precise results
obtained from \empymod. The blue circles indicate the actually computed
responses and the black dots the frequencies which are interpolated or set to
zero. The resulting time-domain response has a relative error of less than
1\,\% everywhere except for very early times.

@@

The shortest offset defines the highest required frequency, and the largest
offset the lowest required frequency. So the more one can restrict the
necessary offset range, the less frequencies are needed. Another important
factor is how to interpolate and extrapolate from the computed frequencies to
the frequencies required for the Fourier transform. For the FFTLog only
extrapolation for higher and lower frequencies is required. The EM response
becomes highly oscillatory for high frequencies, which makes it very hard to
extrapolate the response to frequencies $f>f_\mathrm{max}$. However, if
$f_\mathrm{max}$ is chosen judiciously, the importance of higher frequencies
for the Fourier transform can be neglected and we can set those responses to
zero. The extrapolation of frequencies $f<f_\mathrm{min}$ can be changed to an
interpolation by assuming a zero imaginary response at zero frequency, and we
then use PCHIP to interpolate the missing frequencies (as we work on a
logarithmic scale we cannot choose 0\,Hz, but instead take $10^{-100}$\,Hz). In
the case of the DLF method we also have to interpolate in between the computed
frequencies, for which we found it better to use a cubic spline. As can be seen
from Figures \ref{fig:GUI-FFTLog-lin} and \ref{fig:GUI-DLF-lin}, using the
FFTLog or the DLF with the same actually computed frequencies results in very
similar responses.

@@

In summary, the adaptive gridding takes $f$, $D_\mr{s}$, $\sigma_\mr{src}$,
$\sigma_\mr{ave}$, $n_\delta$, and ranges for $\alpha_\mr{s}$, $\alpha_\mr{c}$,
where we usually fix $\alpha_\mr{s}=1$ or keep it at least below 1.05, and let
$\alpha_\mr{c}$ be anything between $[1, 1.5]$. The minimum cell width
$\Delta_\mr{min}$ can further be restricted by a user-defined range. Given
these inputs the adaptive gridding will search for the smallest possible number
of cells which fulfils these criteria. The multigrid method implemented in
\emg3d puts some constraints on the number of cells, of which the adaptive
gridding takes care (the number of cells have to be powers of two multiplied by
a low prime, e.g., $\{2,3,5\}\cdot2^n$).

@@

Figure \ref{fig:fullspace} (a) shows the frequency-domain result, where the
blue dots are the computed responses and the black dots correspond to the
interpolated values or the values set to zero.
%
\plot*{fullspace}{width=\textwidth}{
  (a) Frequency- and (b) time-domain results for the homogeneous space model.
  The red lines are the analytical solutions, the blue circles are the actually
  computed responses with \emg3d, the black dots are the interpolated
  responses, and the dashed black line the obtained time-domain response. The
  errors are clipped for values smaller than 0.1\,\% and bigger than 100\,\%.
}
%
Most of the computed values stay below a relative error of 1\,\%, our
chosen adaptive gridding only starts to generate considerable errors at higher
frequencies. Figure \ref{fig:fullspace}(b) shows the corresponding time-domain
result, where the dashed black line is the result from \emg3d, on top of the
red line which is the analytical result. The relative error is mostly below
1\,\%, except for early times. However, for practical reasons that is more than
enough. Figure~\ref{fig:fullspace-log} shows the same on a logarithmic scale,
with times up to 10 seconds. It clearly shows that if later times are required,
we would need to adjust our Fourier transform parameters.
%
\plot{fullspace-log}{width=\cwidth}{
  Same as in Figure~\ref{fig:fullspace} (b), but on a logarithmic scale. To
  improve later times we would have to compute lower frequencies; to improve
  earlier times we would have to compute more frequencies per decade to get a
  better resolution.
}
%
Note that for the gridding we chose $n_\delta=12$, which is very dense. This
was necessary because we are relatively close to the source. If the offsets of
interest are larger this factor can be lowered considerably; 3--4 is often
enough.

@@

For the comparison we use the open-source code \simpeg \citep{CAG.15.Cockett}.
\simpeg is a framework for modelling and inversion of geophysical data such as
gravity, magnetics, and CSEM. It has Maxwell's equations implemented in both
the frequency and time domain. As such we can compare our result computed in
the frequency domain followed by a Fourier transform to a result computed
directly in the time domain. A principal difference between \simpeg and \emg3d
is that the former has various direct solvers implemented, whereas the latter
is an iterative multigrid solver. The 3D model is therefore a rather small
example in order to be able to run it on our test machine, as the memory
requirement by the direct solver would otherwise be too high. There are not
many options out there of open-source time-domain 3D codes, \simpeg being the
one we found to be suitable. A step-off response was chosen as this is the
response currently implemented in \simpeg.

The model was discretised with $100\times100\times100\,$m cells in the survey
domain $D_\mr{s}$. For the \simpeg model, 14 cells in $x$-direction and 12
cells in $y$- and $z$-directions were used on both sides with a stretching of
1.3 for the total computation domain $D_\mr{c}$, which yields a mesh of
\num{58344} cells. The time-steps start at 0.1\,s and are: $21\times0.01\,$s,
$23\times0.03\,$s, $21\times0.1\,$s, $23\times0.3\,$s, covering exactly the
desired range of 0.1--10\,s. For \emg3d, the mesh is generated
frequency-dependent as in the previous examples, with a maximum stretching of
$\alpha_\mr{c}=1.5$. This results in meshes between \num{18432} cells for the
highest frequencies and \num{76800} cells for the lowest frequencies. The
required frequencies were obtained by using the FFTLog with five points per
decade, which results in 20 frequencies between 0.001\,Hz and 8\,Hz. The actual
transform was carried out with the 201-point sine-cosine filter from
\cite{GEO.09.Key}.

The results are shown in Figure~\ref{fig:3d-result}: In (a) the 1D background
responses and the relative error using the semi-analytical result obtained from
\empymod, and in (b) the responses including the target.
%
\plot{3d-result}{width=\textwidth}{
  Responses for the model outlined in Figure~\ref{fig:3d-model} using \simpeg
  and \emg3d, and for the layered background also \empymod. In the lower plot
  of (a) the relative error (\%) in comparison to \empymod is shown, and in (b)
  the normalized difference (\%) between the two 3D codes. The results of these
  plots are clipped for values smaller than 0.1\,\% and bigger than 10\,\%.
}
%
The background comparison shows that both 3D codes do an acceptable job having
a relative error of a few percents at most; the result for \emg3d seems to be
better at early times. We cannot compare the errors for the response that
includes the target for lack of an analytical solution. The background model
from \empymod is only included to show that there is a significant response
from the target. We therefore show the normalized difference (ND) between the
two 3D codes as a percentage, where $\mr{ND}=
50|(\simpeg-\emg3d)/(\simpeg+\emg3d)|$. The ND between the two codes is below
1\,\% everywhere except for early times.

Both codes took a little over 7 minutes to compute the two models (single
thread). However, in this particular comparison, the main difference in runtime
is not frequency-domain computation vs. time-domain computation, but iterative
solver vs. direct solver. For a test we doubled the number of cells in each
direction by using 50\,m for the minimum cell width instead of 100\,m, and
taking also half of the stretching, which results then in a mesh of
\num{466752} cells for \simpeg and meshes between \num{102400} and \num{614400}
for \emg3d. Because the memory of our test machine was not large enough for the
direct solver of \simpeg, the last model had to be run on a cluster. In this
scenario, \emg3d was about eight times faster (single thread).

>>>>

\address{
\footnotemark[1]TU Delft,
Building 23,
Stevinweg 1 / PO-box 5048,
2628 CN Delft;\\
\footnotemark[2]Shell Global Solutions International BV,
Grasweg 31, 1031 HW Amsterdam;\\
E-mail: \href{mailto:Dieter@Werthmuller.org}{Dieter@Werthmuller.org};\\
}

@@

%%fakesection ===    ABSTRACT    ===
\begin{abstract}
%
Modelling three-dimensional time-domain controlled-source electromagnetic data
with frequency-domain solvers is an alternative to time-domain solvers. It
requires the computation of many frequencies if standard Fourier transforms are
used, which can make it prohibitively expensive in comparison with a
time-domain computation. The speed of time-domain modelling with
frequency-domain computation is defined by three key points: solver, method and
implementation of the Fourier transformation, and gridding.
%
The faster the frequency-domain solver, the faster time-domain modelling will
be. It is important that the solver is robust over a wide range of frequencies.
A three-dimensional, iterative, frequency-domain multigrid solver satisfies
this requirement, but any other robust solver will do.
%
The method should require as few frequencies as possible while remaining
robust. As the frequency range spans many orders of magnitude, the required
frequencies are ideally equally spaced on a logarithmic scale. The fast
frequency-to-time domain method uses either the digital linear filter method
or the logarithmic fast Fourier transform together with a careful selection of
frequencies and interpolation. This methodology requires typically 15 to 20
frequencies to cover a wide range of offsets in one-dimensional test models.
%
The gridding should be frequency-dependent, which is accomplished by making it
a function of skin depth. Optimising for the least number of required cells for
a given frequency should be combined with optimising for computational speed.
%
Looking carefully at these three \new{key} points results in much smaller
computation times with speedup factors of ten or more over previous results. A
frequency-domain computation of time-domain data can therefore be competitive
to time-domain solvers if the required frequencies and the corresponding grids
are carefully chosen.
%
\\[1cm] \textbf{Keywords}: Controlled source electromagnetics (CSEM); Numerical
modelling; Fourier analysis.
\end{abstract}

@@

Being able to model CSEM data both in the frequency-domain and in the
time-domain can be desirable, as both domains have advantages and
disadvantages. One way to achieve this is to implement Maxwell's equation in
both domains\old{, as it is done, for instance, in \simpeg (Cockett et al.,
2015)}. Another approach is to have Maxwell's equation only implemented in one
domain, and use Fourier transforms to go to the other. However, this approach
can be costly, as many frequencies over a wide range are required to go from
the time domain to the frequency domain, or many times over a wide range for
the opposite direction. We present a methodology which significantly reduces
the amount and range of the required frequencies, translating into a
significant reduction in computation time. The required computation grids for
low frequencies (in our case in the order of 0.001\,Hz) and high frequencies
(around 100\,Hz) are hugely different. Low frequencies can be computed on a
coarser grid, but they require a much larger domain in order to avoid boundary
effects. High frequencies, on the other hand, require denser gridding, but they
are much more limited in reach. An adaptive gridding scheme is therefore
required. We build our approach upon \cite{GEO.07.Plessix}, who presented such
an adaptive gridding for multi-frequency (and multi-source) CSEM modelling.

@@

For the numerical computations we use the open-source (Apache License 2.0) code
\emg3d \citep{JOSS.19.Werthmuller}, a multigrid solver based on
\cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov subspace
solvers or as a solver on its own. The multigrid approach works fine for the
diffusive approximation of Maxwell's equation, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which \old{\emg3d has implemented }known improvements such as
line-relaxation and semicoarsening \citep{ECCFD.06.Jonsthovel} \new{are
implemented}, with a non-standard Cholesky decomposition to speed up the
computation \citep{GEO.08.Mulder}. One of the big advantages of the multigrid
method is that it scales linearly (optimal) with the grid size in both CPU and
RAM usage \citep{B.Springer.20.Mulder}. This makes it feasible to run even big
models on standard computers, without the need for big clusters. All the
examples in this article are run on a laptop with an i7-6600U\,CPU\,@\,2.6\,GHz
(x4) and 16\,GB of memory, using Ubuntu 18.04 and Python 3.7. For the
semi-analytical computations of layered models we use the open-source (Apache
License 2.0) code \empymod \citep{GEO.17.Werthmuller}.

@@

There are three key-parameters that have to be defined: Minimum and maximum
frequency ($f_\mr{min}, f_\mr{max}$) and the number of frequencies per decade
($n_\mr{dec}^f$). Trial-and-error with a 3D code is very time-intensive.
However, a simplified, layered model for the required survey setup and a fast
1D modeller makes it possible to estimate these parameters easily. To this end,
we created a graphical, interactive tool using \empymod, as shown in
Figure~\ref{fig:GUI-FFTLog-lin}.
%
\plot{GUI-FFTLog-lin}{width=\tcwidth}{
  Interactive frequency selection for a user-provided layered model (the shown
  model parameters are described in the text). This example shows the impulse
  response at an offset of 5\,km, for which it uses FFTLog from 0.001\,Hz to
  10\,Hz with five frequencies per decade. Note that the error is clipped for
  values smaller than 0.01\,\% and bigger than 100\,\%.
}
%
The example model is a marine scenario with 1\,km water depth of resistivity
$\rho=0.3\,\Omega$\,m ($\rho = \sigma^{-1}$) and a 100\,m thick target of
100\,$\Omega$\,m at 1\,km below the seafloor in a background of 1\,$\Omega$\,m.
The source is 50\,m above the seafloor, the receiver is on the seafloor, and
the response is the inline $x$-directed $E$-field. The left subplot shows the
imaginary part in the frequency domain and the right plot the corresponding
impulse response in the time domain. The red lines are the \new{semi-analytical
responses}\old{precise results obtained from \empymod}. The blue circles
indicate the actually computed responses and the black dots the frequencies
which are interpolated or set to zero. The resulting time-domain response has a
relative error of less than 1\,\% everywhere except for very early times.

@@

The shortest offset defines the highest required frequency, and the largest
offset the lowest required frequency. So the more one can restrict the
necessary offset range, the \old{less}\new{fewer} frequencies are needed.
Another important factor is how to interpolate and extrapolate from the
computed frequencies to the frequencies required for the Fourier transform. For
the FFTLog only extrapolation for higher and lower frequencies is required. The
EM response becomes highly oscillatory for high frequencies, which makes it
very hard to extrapolate the response to frequencies $f>f_\mathrm{max}$.
However, if $f_\mathrm{max}$ is chosen judiciously, the importance of higher
frequencies for the Fourier transform can be neglected and we can set those
responses to zero. The extrapolation of frequencies $f<f_\mathrm{min}$ can be
changed to an interpolation by assuming a zero imaginary response at zero
frequency, and we then use PCHIP to interpolate the missing frequencies (as we
work on a logarithmic scale we cannot choose 0\,Hz, but instead take
$10^{-100}$\,Hz). In the case of the DLF method we also have to interpolate in
between the computed frequencies, for which we found it better to use a cubic
spline. As can be seen from Figures \ref{fig:GUI-FFTLog-lin} and
\ref{fig:GUI-DLF-lin}, using the FFTLog or the DLF with the same actually
computed frequencies results in very similar responses.

@@

In summary, the adaptive gridding takes $f$, $D_\mr{s}$, $\sigma_\mr{src}$,
$\sigma_\mr{ave}$, $n_\delta$, and ranges for $\alpha_\mr{s}$, $\alpha_\mr{c}$,
where we usually fix $\alpha_\mr{s}=1$ or keep it at least below 1.05, and let
$\alpha_\mr{c}$ be anything between $[1, 1.5]$. The minimum cell width
$\Delta_\mr{min}$ can further be restricted by a user-defined range. Given
these inputs the adaptive gridding will search for the smallest possible number
of cells which fulfils these criteria. The \new{implemented} multigrid method
\old{implemented in \emg3d} puts some constraints on the number of cells, of
which the adaptive gridding takes care (the number of cells have to be powers
of two multiplied by a low prime, e.g., $\{2,3,5\}\cdot2^n$).

@@

Figure \ref{fig:fullspace} (a) shows the frequency-domain result, where the
blue dots are the computed responses and the black dots correspond to the
interpolated values or the values set to zero.
%
\plot*{fullspace}{width=\textwidth}{
  (a) Frequency- and (b) time-domain results for the homogeneous space model.
  The red lines are the analytical solutions, the blue circles are the actually
  computed responses\old{ with \emg3d}, the black dots are the interpolated
  responses, and the dashed black line the obtained time-domain response. The
  errors are clipped for values smaller than 0.1\,\% and bigger than 100\,\%.
}
%
Most of the computed values stay below a relative error of 1\,\%, our chosen
adaptive gridding only starts to generate considerable errors at higher
frequencies. Figure \ref{fig:fullspace}(b) shows the corresponding time-domain
result, where the dashed black line is the result \old{from
\emg3d}\new{obtained by transformation of the frequency-domain response}, on
top of the red line which is the analytical result. The relative error is
mostly below 1\,\%, except for early times. However, for practical reasons that
is more than enough. Figure~\ref{fig:fullspace-log} shows the same on a
logarithmic scale, with times up to 10 seconds. It clearly shows that if later
times are required, we would need to adjust our Fourier transform parameters.
%
\plot{fullspace-log}{width=\cwidth}{
  Same as in Figure~\ref{fig:fullspace} (b), but on a logarithmic scale. To
  improve later times we would have to compute lower frequencies; to improve
  earlier times we would have to compute more frequencies per decade to get a
  better resolution.
}
%
Note that for the gridding we chose $n_\delta=12$, which is very dense. This
was necessary because we are relatively close to the source. If the offsets of
interest are larger this factor can be lowered considerably; 3--4 is often
enough.

@@

For the comparison we use the open-source code \simpeg
\citep{CAG.15.Cockett}\old{. \simpeg}\new{, which} is a framework for modelling
and inversion of geophysical data such as gravity, magnetics, and CSEM. It has
Maxwell's equations implemented in both the frequency and time domain. As such
we can compare our result computed in the frequency domain followed by a
Fourier transform to a result computed directly in the time domain. A principal
difference between \simpeg and \emg3d is that the former has various direct
solvers implemented, whereas the latter is an iterative multigrid solver. The
3D model is therefore a rather small example in order to be able to run it on
our test machine, as the memory requirement by the direct solver would
otherwise be too high. There are not many options out there of open-source
time-domain 3D codes, \simpeg being the one we found to be suitable. A step-off
response was chosen as this is the response currently implemented in \new{the
framework}\old{\simpeg}.

The model was discretised with $100\times100\times100\,$m cells in the survey
domain $D_\mr{s}$. For the \old{\simpeg}\new{time-domain} model, 14 cells in
$x$-direction and 12 cells in $y$- and $z$-directions were used on both sides
with a stretching of 1.3 for the total computation domain $D_\mr{c}$, which
yields a mesh of \num{58344} cells. The time-steps start at 0.1\,s and are:
$21\times0.01\,$s, $23\times0.03\,$s, $21\times0.1\,$s, $23\times0.3\,$s,
covering exactly the desired range of 0.1--10\,s. For
\old{\emg3d}\new{frequency-domain model}, the mesh is generated
frequency-dependent as in the previous examples, with a maximum stretching of
$\alpha_\mr{c}=1.5$. This results in meshes between \num{18432} cells for the
highest frequencies and \num{76800} cells for the lowest frequencies. The
required frequencies were obtained by using the FFTLog with five points per
decade, which results in 20 frequencies between 0.001\,Hz and 8\,Hz. The actual
transform was carried out with the 201-point sine-cosine filter from
\cite{GEO.09.Key}.

The results are shown in Figure~\ref{fig:3d-result}: In (a) the 1D background
responses and the relative error using the semi-analytical result\old{obtained
from \empymod}, and in (b) the responses including the target.
%
\plot{3d-result}{width=\textwidth}{
  Responses for the model outlined in Figure~\ref{fig:3d-model} using
  \new{time-domain}\old{\simpeg} and \new{frequency-domain
  computations}\old{\emg3d}, and for the layered background also \new{the
  semi-analytical result}\old{\empymod}. In the lower plot of (a) the relative
  error (\%) \new{is shown} in comparison to \new{the semi-analytical
  result}\old{\empymod is shown}, and in (b) the normalized difference (\%)
  between the two 3D codes. The results of these plots are clipped for values
  smaller than 0.1\,\% and bigger than 10\,\%.
}
%
The background comparison shows that both 3D codes do an acceptable job having
a relative error of a few percents at most; the result \new{obtained through
transformation}\old{for \emg3d} seems to be better at early times. We cannot
compare the errors for the response that includes the target for lack of an
analytical solution. The \new{semi-analytical} background model \old{from
\empymod }is only included to show that there is a significant response from
the target. We therefore show the normalized difference (\old{ND}\new{NRMSD})
between the two \old{3D codes}\new{responses $R_1$ and $R_2$} as a percentage,
where \old{$\mr{ND}= 50|(\simpeg-\emg3d)/(\simpeg+\emg3d)|$.}
\new{$\mr{NRMSD~(\%)} = 200|R_1 - R_2|/(|R_1| + |R_2|)$.} The
\old{ND}\new{NRMSD} between the two codes is below 1\,\% everywhere except for
early times.

Both codes took a little over 7 minutes to compute the two models (single
thread). However, in this particular comparison, the main difference in runtime
is not frequency-domain computation vs. time-domain computation, but iterative
solver vs. direct solver. For a test we doubled the number of cells in each
direction by using 50\,m for the minimum cell width instead of 100\,m, and
taking also half of the stretching, which results then in a mesh of
\num{466752} cells \old{for \simpeg}\new{in the time domain} and meshes between
\num{102400} and \num{614400} \old{for \emg3d}\new{in the frequency domain}.
Because the memory of our test machine was not large enough for the direct
solver of \simpeg, the last model had to be run on a cluster. In this scenario,
\emg3d was about eight times faster (single thread).
